---
title: "What's the Half-Life of the Economic Vote? (About a Year and a Half)"
author: "Jack Bailey"
date: "27 October 2020"
output:
  xaringan::moon_reader:
    css: "xaringan-jb.css"
    lib_dir: libs
    nature:
      navigation:
        scroll: false
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo = F, include = F, cache = T}

# Load packages

library(xaringanthemer)
library(kableExtra)
library(tidyverse)
library(magrittr)
library(jbmisc) # https://github.com/jackobailey/jbmisc
library(brms)
library(here)


# Load half-life model

m1 <- readRDS(here("_output", "m1.rds"))


# Estimate half-life distribution

half_life_dist <-
  log(2) %>% 
  divide_by(
    posterior_samples(
      m1,
      pars = "b_lambda"
    ) %>% 
      pluck(1)
  )


# Get posterior samples for easy access

post_samps <- 
  posterior_samples(m1) %>% 
  tibble() %>% 
  select(
    beta0 = b_beta0_Intercept,
    lambda = b_lambda_Intercept
  )


# Load second robustness model

m3 <- readRDS(here("_output", "m3.rds"))


# Get posterior economic voting effect from model 3

m3_econ <- 
  posterior_samples(m3, pars = "gdp") %>% 
  tibble()


# Load figure scripts

source(here("_scripts", "003_gdp_plot.R"))
source(here("_scripts", "004_decay_plot.R"))
source(here("_scripts", "005_slope_plot.R"))
source(here("_scripts", "006_real_decay_plot.R"))
source(here("_scripts", "007_real_slope_plot.R"))
source(here("_scripts", "008_table.R"))


# Load raw GDP data

gdp <- read_csv(here("_data", "gdp.csv"))


```


class: inverse, center, middle

## "After decades of attention, we are little beyond introspection in understanding the processes by which citizens come to perceive economic movement"

.right[*MacKuen et al. (1992, p. 597)*]

---

## Voter Myopia

--

### Our Understanding is Limited

--

Often rely on ad-hoc assumptions

--

And because different scholars make different assumptions, the time frames that they assume voters to use differ too.

--

### So What?

--

Myopic voters are open to abuse

--

1. Voter myopia lets governments do what they like early in their term, without any meaningful consequences

--

2. Myopic voters might vote not for the best economic *managers*, but instead the best economic *manipulators*

---
class: inverse, center, middle

# How Retrospective are<br>Retrospective Voters?

---

## Voter-centric Research

--

### Steirs et al. (2019)

--

Voters are "attentive to the government's performance in the long run as well as the short" (p. 647)

--

Model incumbent support as function of past government satisfaction and economic satisfaction

--

But their models focus on *levels* and not *change in levels* so tell us little about retrospective vote

--

### Huber et al. (2012) and Healy & Lenz (2014)

--

Voters use heuristics: peak-end rule

--

Voters intend to judge based on cumulative performance, but instead "substitute the end for the whole"

---

## Electorate-centric Research

--

### Hibbs (1987)

--

Examines both if electorates are myopic and what functional form their myopia takes

--

Assumes the effect of economic change decays at some known exponential rate

--

Uses this to weight year-on-year economic change for each quarter in post-war each presidential term

--

### Wlezien (2015)

--

Similar approach to Hibbs, but more elections

--

Argues for a logistic, not exponential decay function

---

## A Hybrid Approach

--

Both approaches have problems

--

**Voter-centric:** Economic perception items suffer from serious shortcomings (Bailey, 2019; Bisgaard, 2019; Evans & Pickup, 2010)

--

**Electorate-centric:** Almost impossible to overcome the problem of ecological inference (Stewart & Clarke, 2017)

--

Why not combine the best of both approaches?

--

**Individual-level vote choice:** Avoids having to deal with problems of ecological inference

--

**Aggregate-level economic statistics:** Avoids any systematic perceptual biases that leak into reported perceptions

---

class: inverse, center, middle

# Data and Methods

---

## Data

--

### Individual-level Data:

--

BES Continuous Monitoring Survey (Whiteley et al., 2014)

--

Monthly cross-sections, April 2004 to February 2014

--

132,369 people took part

--

DV: Vote choice (1 = Incumbent, 0 = Other)

--

### Aggregate-level Data

--

Monthly UK GDP estimates from the ONS

--

Characterised by growth, but also the global financial crisis

---

## Methods

--

I take a two-staged approach:

--

### Stage 1: Estimate Fine-Grained GDP

--

Aggregate data impose a major constraint: data is structured according to the release schedule of the statistical agencies behind them

--

### Stage 2: Estimate Voter Myopia Model

---
class: inverse, center, middle

# Stage 1: Estimating Daily GDP

---

class: center, middle

```{r gdp-panel-1, fig.height = 3, fig.width = 5, dpi = 500, echo = F, cache = T}
panel1 + theme_bailey() + theme(axis.title.x = element_blank())
```

---

class: center, middle

```{r gdp-panel-2, fig.height = 3, fig.width = 5, dpi = 500, echo = F, cache = T}
panel2 + theme_bailey() + theme(axis.title.x = element_blank())
```

---

class: center, middle

```{r gdp-panel-3, fig.height = 3, fig.width = 5, dpi = 500, echo = F, cache = T}
panel3 + theme_bailey() + theme(axis.title.x = element_blank())
```

---
class: inverse, center, middle

# Stage 2: Building the Model

---

class: middle

.pull-left[
```{r eq-1, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    gdp = seq(-5, 5, by = 0.1),
    vote = inv_logit(0 + betaT(time = 0, beta0 = 0.2, lambda = 1)*gdp)
) %>% 
  ggplot(
    aes(
      y = vote,
      x = gdp
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(
    breaks = seq(-5, 5, by = 2.5),
    labels = c("-5%", "-2.5%", "0%", "+2.5%", "+5%")
  ) +
  labs(
    y = "Prob. Vote Incumbent",
    x = "Change in Conditions"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = bailey_colours("grey6")
    ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha + \beta (C_{0} - C_{t})$

]

---

class: middle

.pull-left[
```{r eq-2, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    gdp = seq(-5, 5, by = 0.1),
    vote = inv_logit(0 + betaT(time = 0, beta0 = 0.2, lambda = 1)*gdp)
) %>% 
  ggplot(
    aes(
      y = vote,
      x = gdp
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(
    breaks = seq(-5, 5, by = 2.5),
    labels = c("-5%", "-2.5%", "0%", "+2.5%", "+5%")
  ) +
  labs(
    y = "Prob. Vote Incumbent",
    x = "Change in Conditions"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = bailey_colours("grey6")
    ) +
  geom_line(lineend = "round") +
  geom_segment(
    x = 0,
    xend = 0,
    y = 0.55,
    yend = 0.70,
    lineend = "round",
    size = 0.5,
    color = bailey_colours("red"),
    arrow = arrow(ends = "last", type = "closed", length = unit(1.5, "pt"))) +
  geom_segment(
    x = 0,
    xend = 0,
    y = 0.45,
    yend = 0.30,
    lineend = "round",
    size = 0.5,
    color = bailey_colours("red"),
    arrow = arrow(ends = "last", type = "closed", length = unit(1.5, "pt"))) +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta (C_{0} - C_{t})$

]

---

class: middle

.pull-left[
```{r eq-3, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    gdp = seq(-5, 5, by = 0.1),
    vote = inv_logit(0 + betaT(time = 0, beta0 = 0.2, lambda = 1)*gdp)
) %>% 
  ggplot(
    aes(
      y = vote,
      x = gdp
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(
    breaks = seq(-5, 5, by = 2.5),
    labels = c("-5%", "-2.5%", "0%", "+2.5%", "+5%")
  ) +
  labs(
    y = "Prob. Vote Incumbent",
    x = "Change in GDP"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = bailey_colours("grey6")
    ) +
  geom_line(lineend = "round") +
  geom_segment(
    x = 0,
    xend = 0,
    y = 0.55,
    yend = 0.70,
    lineend = "round",
    size = 0.5,
    color = bailey_colours("red"),
    arrow = arrow(ends = "last", type = "closed", length = unit(1.5, "pt"))) +
  geom_segment(
    x = 0,
    xend = 0,
    y = 0.45,
    yend = 0.30,
    lineend = "round",
    size = 0.5,
    color = bailey_colours("red"),
    arrow = arrow(ends = "last", type = "closed", length = unit(1.5, "pt"))) +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta \Delta GDP_{t}$

]


---

class: middle

.pull-left[
```{r eq-4, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    time = seq(0, 5, by = .1),
    beta = 1
) %>% 
  ggplot(
    aes(
      y = beta,
      x = time
    )
  ) +
  scale_y_continuous(breaks = c(0, .5, 1), labels = c("0B", "½B", "1B")) +
  scale_x_continuous(breaks = c(0, 2.5, 5), labels = c("0", "2.5", "5")) +
  coord_cartesian(ylim = c(0, 1), clip = "off") +
  labs(
    y = "Parameter Value",
    x = "Economic Time Frame (Years)"
  ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta$

]


---

class: middle

.pull-left[
```{r eq-5, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    time = seq(0, 5, by = .1),
    beta = betaT(time = time, beta0 = 1, lambda = 0.8)
) %>% 
  ggplot(
    aes(
      y = beta,
      x = time
    )
  ) +
  scale_y_continuous(breaks = c(0, .5, 1), labels = c("0B", "½B", "1B")) +
  scale_x_continuous(breaks = c(0, 2.5, 5), labels = c("0", "2.5", "5")) +
  coord_cartesian(ylim = c(0, 1), clip = "off") +
  labs(
    y = "Parameter Value",
    x = "Economic Time Frame (Years)"
  ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

]


---

class: middle

.pull-left[
```{r eq-6, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

decay = 0.80

labs_h <-
  tibble(
    x = 0,
    xend = log(2)/decay,
    y = .5,
    yend = .5,
    label = paste0("λ = ", decay)
  )

labs_v <-
  tibble(
    x = log(2)/decay,
    xend = log(2)/decay,
    y = 0,
    yend = .5,
    label = paste0("λ = ", decay)
  )

tibble(
    time = seq(0, 5, by = .1),
    beta = betaT(time = time, beta0 = 1, lambda = decay)
) %>% 
  ggplot(
    aes(
      y = beta,
      x = time
    )
  ) +
  scale_y_continuous(breaks = c(0, .5, 1), labels = c("0B", "½B", "1B")) +
  scale_x_continuous(breaks = c(0, 5), labels = c("0", "5")) +
  coord_cartesian(ylim = c(0, 1), clip = "off") +
  labs(
    y = "Parameter Value",
    x = "Economic Time Frame (Years)"
  ) +
  geom_segment(
    data = labs_h,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("white"),
    lineend = "round"
  ) +
  geom_segment(
    data = labs_h,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("red"),
    linetype = "dotted",
    lineend = "round"
  ) +
  geom_segment(
    data = labs_v,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("red"),
    linetype = "dotted",
    lineend = "round"
  ) +
  geom_segment(
    data = labs_v,
    aes(
      x = x,
      xend = xend,
      y = -.05,
      yend = -.08
    ),
    colour = bailey_colours("black"),
    lineend = "round"
  ) +
  geom_text(
    data = labs_v,
    aes(
      x = x,
      y = y-.12,
      label = paste("t½ =", round(xend, 1))
    ),
    colour = bailey_colours("black"),
    family = "Cabin",
    size = 2.3
  ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

$t_{1/2} = \frac{log(2)}{\lambda}$
]

---

class: middle

.pull-left[
```{r eq-7, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

decay = 0.6

labs_h <-
  tibble(
    x = 0,
    xend = log(2)/decay,
    y = .5,
    yend = .5,
    label = paste0("λ = ", decay)
  )

labs_v <-
  tibble(
    x = log(2)/decay,
    xend = log(2)/decay,
    y = 0,
    yend = .5,
    label = paste0("λ = ", decay)
  )

tibble(
    time = seq(0, 5, by = .1),
    beta = betaT(time = time, beta0 = 1, lambda = decay)
) %>% 
  ggplot(
    aes(
      y = beta,
      x = time
    )
  ) +
  scale_y_continuous(breaks = c(0, .5, 1), labels = c("0B", "½B", "1B")) +
  scale_x_continuous(breaks = c(0, 5), labels = c("0", "5")) +
  coord_cartesian(ylim = c(0, 1), clip = "off") +
  labs(
    y = "Parameter Value",
    x = "Economic Time Frame (Years)"
  ) +
  geom_segment(
    data = labs_h,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("white"),
    lineend = "round"
  ) +
  geom_segment(
    data = labs_h,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("red"),
    linetype = "dotted",
    lineend = "round"
  ) +
  geom_segment(
    data = labs_v,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("red"),
    linetype = "dotted",
    lineend = "round"
  ) +
  geom_segment(
    data = labs_v,
    aes(
      x = x,
      xend = xend,
      y = -.05,
      yend = -.08
    ),
    colour = bailey_colours("black"),
    lineend = "round"
  ) +
  geom_text(
    data = labs_v,
    aes(
      x = x,
      y = y-.12,
      label = paste("t½ =", round(xend, 1))
    ),
    colour = bailey_colours("black"),
    family = "Cabin",
    size = 2.3
  ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

$t_{1/2} = \frac{log(2)}{\lambda}$
]


---

class: middle

.pull-left[
```{r eq-8, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

decay = 0.4

labs_h <-
  tibble(
    x = 0,
    xend = log(2)/decay,
    y = .5,
    yend = .5,
    label = paste0("λ = ", decay)
  )

labs_v <-
  tibble(
    x = log(2)/decay,
    xend = log(2)/decay,
    y = 0,
    yend = .5,
    label = paste0("λ = ", decay)
  )

tibble(
    time = seq(0, 5, by = .1),
    beta = betaT(time = time, beta0 = 1, lambda = decay)
) %>% 
  ggplot(
    aes(
      y = beta,
      x = time
    )
  ) +
  scale_y_continuous(breaks = c(0, .5, 1), labels = c("0B", "½B", "1B")) +
  scale_x_continuous(breaks = c(0, 5), labels = c("0", "5")) +
  coord_cartesian(ylim = c(0, 1), clip = "off") +
  labs(
    y = "Parameter Value",
    x = "Economic Time Frame (Years)"
  ) +
  geom_segment(
    data = labs_h,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("white"),
    lineend = "round"
  ) +
  geom_segment(
    data = labs_h,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("red"),
    linetype = "dotted",
    lineend = "round"
  ) +
  geom_segment(
    data = labs_v,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("red"),
    linetype = "dotted",
    lineend = "round"
  ) +
  geom_segment(
    data = labs_v,
    aes(
      x = x,
      xend = xend,
      y = -.05,
      yend = -.08
    ),
    colour = bailey_colours("black"),
    lineend = "round"
  ) +
  geom_text(
    data = labs_v,
    aes(
      x = x,
      y = y-.12,
      label = paste("t½ =", round(xend, 1))
    ),
    colour = bailey_colours("black"),
    family = "Cabin",
    size = 2.3
  ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

$t_{1/2} = \frac{log(2)}{\lambda}$
]


---

class: middle

.pull-left[
```{r eq-9, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

decay = 0.2

labs_h <-
  tibble(
    x = 0,
    xend = log(2)/decay,
    y = .5,
    yend = .5,
    label = paste0("λ = ", decay)
  )

labs_v <-
  tibble(
    x = log(2)/decay,
    xend = log(2)/decay,
    y = 0,
    yend = .5,
    label = paste0("λ = ", decay)
  )

tibble(
    time = seq(0, 5, by = .1),
    beta = betaT(time = time, beta0 = 1, lambda = decay)
) %>% 
  ggplot(
    aes(
      y = beta,
      x = time
    )
  ) +
  scale_y_continuous(breaks = c(0, .5, 1), labels = c("0B", "½B", "1B")) +
  scale_x_continuous(breaks = c(0, 5), labels = c("0", "5")) +
  coord_cartesian(ylim = c(0, 1), clip = "off") +
  labs(
    y = "Parameter Value",
    x = "Economic Time Frame (Years)"
  ) +
  geom_segment(
    data = labs_h,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("white"),
    lineend = "round"
  ) +
  geom_segment(
    data = labs_h,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("red"),
    linetype = "dotted",
    lineend = "round"
  ) +
  geom_segment(
    data = labs_v,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    colour = bailey_colours("red"),
    linetype = "dotted",
    lineend = "round"
  ) +
  geom_segment(
    data = labs_v,
    aes(
      x = x,
      xend = xend,
      y = -.05,
      yend = -.08
    ),
    colour = bailey_colours("black"),
    lineend = "round"
  ) +
  geom_text(
    data = labs_v,
    aes(
      x = x,
      y = y-.12,
      label = paste("t½ =", round(xend, 1))
    ),
    colour = bailey_colours("black"),
    family = "Cabin",
    size = 2.3
  ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

$t_{1/2} = \frac{log(2)}{\lambda}$
]


---

class: middle

.pull-left[
```{r eq-10, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    gdp = seq(-5, 5, by = 0.1),
    vote = inv_logit(0 + betaT(time = 0, beta0 = 0.2, lambda = 1)*gdp)
) %>% 
  ggplot(
    aes(
      y = vote,
      x = gdp
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(
    breaks = seq(-5, 5, by = 2.5),
    labels = c("-5%", "-2.5%", "0%", "+2.5%", "+5%")
  ) +
  labs(
    y = "Prob. Vote Incumbent",
    x = "Change in Conditions"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = bailey_colours("grey6")
    ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

$t_{1/2} = \frac{log(2)}{\lambda}$
]

---

class: middle

.pull-left[
```{r eq-11, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    gdp = seq(-5, 5, by = 0.1),
    vote = inv_logit(0 + betaT(time = 0.5, beta0 = 0.2, lambda = 1)*gdp)
) %>% 
  ggplot(
    aes(
      y = vote,
      x = gdp
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(
    breaks = seq(-5, 5, by = 2.5),
    labels = c("-5%", "-2.5%", "0%", "+2.5%", "+5%")
  ) +
  labs(
    y = "Prob. Vote Incumbent",
    x = "Change in Conditions"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = bailey_colours("grey6")
    ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

$t_{1/2} = \frac{log(2)}{\lambda}$
]

---

class: middle

.pull-left[
```{r eq-12, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    gdp = seq(-5, 5, by = 0.1),
    vote = inv_logit(0 + betaT(time = 1, beta0 = 0.2, lambda = 1)*gdp)
) %>% 
  ggplot(
    aes(
      y = vote,
      x = gdp
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(
    breaks = seq(-5, 5, by = 2.5),
    labels = c("-5%", "-2.5%", "0%", "+2.5%", "+5%")
  ) +
  labs(
    y = "Prob. Vote Incumbent",
    x = "Change in Conditions"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = bailey_colours("grey6")
    ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

$t_{1/2} = \frac{log(2)}{\lambda}$
]

---

class: middle

.pull-left[
```{r eq-13, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    gdp = seq(-5, 5, by = 0.1),
    vote = inv_logit(0 + betaT(time = 1.5, beta0 = 0.2, lambda = 1)*gdp)
) %>% 
  ggplot(
    aes(
      y = vote,
      x = gdp
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(
    breaks = seq(-5, 5, by = 2.5),
    labels = c("-5%", "-2.5%", "0%", "+2.5%", "+5%")
  ) +
  labs(
    y = "Prob. Vote Incumbent",
    x = "Change in Conditions"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = bailey_colours("grey6")
    ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

$t_{1/2} = \frac{log(2)}{\lambda}$
]

---

class: middle

.pull-left[
```{r eq-14, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    gdp = seq(-5, 5, by = 0.1),
    vote = inv_logit(0 + betaT(time = 2, beta0 = 0.2, lambda = 1)*gdp)
) %>% 
  ggplot(
    aes(
      y = vote,
      x = gdp
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(
    breaks = seq(-5, 5, by = 2.5),
    labels = c("-5%", "-2.5%", "0%", "+2.5%", "+5%")
  ) +
  labs(
    y = "Prob. Vote Incumbent",
    x = "Change in Conditions"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = bailey_colours("grey6")
    ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

$t_{1/2} = \frac{log(2)}{\lambda}$
]

---

class: middle

.pull-left[
```{r eq-15, fig.asp = 1, fig.width = 3, dpi = 500, echo = F, cache = T}

betaT <- function(time = 0, beta0 = 1, lambda = 1){
  beta0*exp(-lambda*time)
}

tibble(
    gdp = seq(-5, 5, by = 0.1),
    vote = inv_logit(0 + betaT(time = 2.5, beta0 = 0.2, lambda = 1)*gdp)
) %>% 
  ggplot(
    aes(
      y = vote,
      x = gdp
    )
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_x_continuous(
    breaks = seq(-5, 5, by = 2.5),
    labels = c("-5%", "-2.5%", "0%", "+2.5%", "+5%")
  ) +
  labs(
    y = "Prob. Vote Incumbent",
    x = "Change in Conditions"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = bailey_colours("grey6")
    ) +
  geom_line(lineend = "round") +
  theme_bailey() +
  theme(
    axis.ticks.y = element_line(lineend = "round")
  )

```
]

.pull-right[

$\mathrm{ }$

$Vote_{i} \sim \mathrm{Bernoulli}(\pi_{i})$

$logit(\pi_{i}) = \alpha_{wave[i]} + \beta_{t} \Delta GDP_{t}$

$\beta_{t} = \beta_{0}e^{-\lambda t}$

$t_{1/2} = \frac{log(2)}{\lambda}$
]

---
class: inverse, center, middle

# Results

---

class: center, middle

```{r table, echo = F, cache = T}


# Get each draw from the posterior distribution

draws <- posterior_samples(m1)


# Get regression coefficients

draws <-
  draws %>% 
  select(
    Intercept = b_pars_Intercept,
    `GDP (t = 0)` = b_beta0_Intercept,
    `Decay Constant` = b_lambda_Intercept,
    `Years Passed` = b_pars_year,
    `Time in Office` = b_pars_office,
    `Gordon Brown (vs. Tony Blair)` = b_pars_leaderGordonBrown,
    `David Cameron (vs. Tony Blair)` = b_pars_leaderDavidCameron,
    `Gordon Brown x Time` = `b_pars_office:leaderGordonBrown`,
    `David Cameron x Time` = `b_pars_office:leaderDavidCameron`
  )


# Pivot table to long-format then compute summary statistics

table <- 
  draws %>% 
  pivot_longer(cols = everything()) %>% 
  mutate(
    name =
      name %>% 
      factor(
        levels =
          c(
            "Intercept",
            "GDP (t = 0)",
            "Decay Constant",
            "Years Passed",
            "Time in Office",
            "Gordon Brown (vs. Tony Blair)",
            "David Cameron (vs. Tony Blair)",
            "Gordon Brown x Time",
            "David Cameron x Time"
          )
      )
  ) %>% 
  group_by(name) %>% 
  summarise(
    Median = format(round(median(value), 2), nsmall = 2),
    Error = format(round(sd(value), 2), nsmall = 2),
    `2.5%` = format(round(quantile(value, probs = 0.025), 2), nsmall = 2),
    `97.5%` = format(round(quantile(value, probs = 0.975), 2), nsmall = 2),
    .groups = "drop"
  ) %>% 
  rename(
    " " = name
  )


# Create latex table

table %>% 
  kable(
    format = "html",
    align = "lcccc"
    ) %>% 
  kable_styling(
    html_font = "Cabin"
  )

```

---

class: center, middle

```{r real-decay-plot, fig.height = 3.5, fig.width = 6, dpi = 500, echo = F, message = F, cache = T}
real_decay_plot +
  scale_y_continuous(
    breaks = c(0, median(pars$beta0)/2, median(pars$beta0), median(pars$beta0)*2),
    labels = c(0, round(median(pars$beta0) * c(.5, 1, 2), 2)),
    sec.axis = 
      dup_axis(
        trans = ~.,
        labels = c("0B", "½B", "1B", "2B")
      )
  )
```

---

class: center, middle

```{r real-slope-plot, fig.height = 2.5, fig.width = 6, dpi = 500, echo = F, cache = T}
real_slope_plot
```

---

class: center, middle

```{r table-std, echo = F, cache = T}


# Load model 2

m2 <- readRDS(here("_output", "m2.rds"))


# Get each draw from the posterior distribution

draws <- posterior_samples(m2)


# Get regression coefficients

draws <-
  draws %>% 
  select(
    Intercept = b_Intercept,
    `Year-on-Year GDP Change` = b_gdp,
    `Years Passed` = b_year,
    `Time in Office` = b_office,
    `Gordon Brown (vs. Tony Blair)` = b_leaderGordonBrown,
    `David Cameron (vs. Tony Blair)` = b_leaderDavidCameron,
    `Gordon Brown x Time` = `b_office:leaderGordonBrown`,
    `David Cameron x Time` = `b_office:leaderDavidCameron`
  )


# Pivot table to long-format then compute summary statistics

table <- 
  draws %>% 
  pivot_longer(cols = everything()) %>% 
  mutate(
    name =
      name %>% 
      factor(
        levels =
          c(
            "Intercept",
            "Year-on-Year GDP Change",
            "Years Passed",
            "Time in Office",
            "Gordon Brown (vs. Tony Blair)",
            "David Cameron (vs. Tony Blair)",
            "Gordon Brown x Time",
            "David Cameron x Time"
          )
      )
  ) %>% 
  group_by(name) %>% 
  summarise(
    Median = format(round(median(value), 2), nsmall = 2),
    Error = format(round(sd(value), 2), nsmall = 2),
    `2.5%` = format(round(quantile(value, probs = 0.025), 2), nsmall = 2),
    `97.5%` = format(round(quantile(value, probs = 0.975), 2), nsmall = 2),
    .groups = "drop"
  ) %>% 
  rename(
    " " = name
  )


# Create latex table

table %>% 
  kable(
    format = "html",
    align = "lcccc"
    ) %>% 
  kable_styling(
    html_font = "Cabin"
  )

```

---

class: center, middle

```{r table-hl, echo = F, cache = T}


# Load model 2

m3 <- readRDS(here("_output", "m3.rds"))


# Get each draw from the posterior distribution

draws <- posterior_samples(m3)


# Get regression coefficients

draws <-
  draws %>% 
  select(
    Intercept = b_Intercept,
    `GDP Change (t = 1.58)` = b_gdp,
    `Years Passed` = b_year,
    `Time in Office` = b_office,
    `Gordon Brown (vs. Tony Blair)` = b_leaderGordonBrown,
    `David Cameron (vs. Tony Blair)` = b_leaderDavidCameron,
    `Gordon Brown x Time` = `b_office:leaderGordonBrown`,
    `David Cameron x Time` = `b_office:leaderDavidCameron`
  )


# Pivot table to long-format then compute summary statistics

table <- 
  draws %>% 
  pivot_longer(cols = everything()) %>% 
  mutate(
    name =
      name %>% 
      factor(
        levels =
          c(
            "Intercept",
            "GDP Change (t = 1.58)",
            "Years Passed",
            "Time in Office",
            "Gordon Brown (vs. Tony Blair)",
            "David Cameron (vs. Tony Blair)",
            "Gordon Brown x Time",
            "David Cameron x Time"
          )
      )
  ) %>% 
  group_by(name) %>% 
  summarise(
    Median = format(round(median(value), 2), nsmall = 2),
    Error = format(round(sd(value), 2), nsmall = 2),
    `2.5%` = format(round(quantile(value, probs = 0.025), 2), nsmall = 2),
    `97.5%` = format(round(quantile(value, probs = 0.975), 2), nsmall = 2),
    .groups = "drop"
  ) %>% 
  rename(
    " " = name
  )


# Create latex table

table %>% 
  kable(
    format = "html",
    align = "lcccc"
    ) %>% 
  kable_styling(
    html_font = "Cabin"
  )

```

---

## Voter Myopia is Real

--

Voters are myopic and respond most strongly to economic change in the recent past

--

Still, they are not quite as short-sighted as we thought

--

They don't consider only the election year (Healy and Lenz, 2014) or even just the past few months (Achen & Bartels, 2016)

--

But neither do state of the economy several years ago have some "abiding" effect on how they vote (Stiers et al., 2019)

--

Instead, they respond to economic change over the past few years, but with diminishing effect

---

## Implications

--

Might be reassuring: Governments must contend with the consequences of more of their actions

--

Yet the two issues I raised at the start remain a problem:

--

1. "Myopic policies for myopic voters" are still a concern

--

2. Voters are myopic enough that they might favour "manipulators" over "managers"

--

Still possible that economic change influences voters through other means: change in issue ownership or perceptions of competence (Fieldhouse et al., 2020)

--

Lots of potential for future research: grievance asymmetry, comparative perspective, model the decay rate

---

class: center, middle

# Thank You

[**jack-bailey.co.uk**](https://www.jack-bailey.co.uk)

[**jack.bailey@manchester.ac.uk**](mailto:jack.bailey@manchester.ac.uk)

[**@PoliSciJack**](https://www.twitter.com/PoliSciJack)

